---
title: "Data Exploration"
author: "Breanna Jones, Alex Johnson, Lily Foote"
date: "3/4/2024"
output: html_document
---

## Packages Used

```{r packages, message = FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
```

## Reading in Data

```{r reading}
olym <- read.csv(file = "dataset_olympics.csv")
region <- read.csv(file = "noc_region.csv")
```

## Creating data "subsets"

```{r subsets1}
olym$NOC <- gsub("SGP", "SIN", olym$NOC)

olym_sport <- olym %>% 
  rename(Discipline = Sport) %>% 
  mutate(Sport = case_when(
    Discipline == "Swimming" | 
      Discipline == "Diving" |
      Discipline == "Water Polo" |
      Discipline == "Synchronized Swimming" ~ "Aquatics",
    Discipline == "Volleyball" |
      Discipline == "Beach Volleyball" ~ "Volleyball",
    Discipline == "Gymnastics" |
      Discipline == "Rhythmic Gymnastics" ~ "Gymnastics",
    Discipline == "Bobsleigh" |
      Discipline == "Skeleton" ~ "Bobsleigh",
    Discipline == "Speed Skating" |
      Discipline == "Short Track Speed Skating" |
      Discipline == "Figure Skating" ~ "Skating",
    Discipline == "Alpine Skiing" |
      Discipline == "Cross Country Skiing" |
      Discipline == "Ski Jumping" |
      Discipline == "Nordic Combined" |
      Discipline == "Freestyle Skiing" |
      Discipline == "Snowboarding" ~ "Skiing")) %>% 
  filter(Sport == "Skiing" | Sport == "Skating" | Sport == "Bobsleigh" | Sport == "Gymnastics" | Sport == "Volleyball" | Sport == "Aquatics") %>%
  filter(Year >= 1980) %>% 
  na.omit() %>% 
  mutate(is_medalist = ifelse((Medal == "Bronze" | Medal == "Silver" | Medal == "Gold"), "medalist", "non-medalist"))
```

```{r subsets2}
olym_reg <- left_join(olym_sport, region, by = join_by(NOC == noc_region))
olym_reg <- na.omit(olym_reg)

olym_reg_medals <- olym_reg %>% 
  filter(is_medalist == "medalist")

```

## Cleaning the data

```{r clean-data}
olym_clean <- olym_reg %>%
  rename(Region = reg) %>% #for continuity 
  select(ID, Name, Sex, Age, Height, Weight, Region, NOC, Year, Season, City, Sport, Discipline, Event, is_medalist, Medal) #reordering cols
```


## First look

```{r glimpse}
glimpse(olym_reg)
glimpse(olym_reg_medals)

olym_sport %>% 
  group_by(Sport) %>% 
  summarize(count = n(), avg_age = mean(Age), med_age = median(Age), sd_age = sd(Age), IQR_age = IQR(Age), max_age = max(Age), min_age = min(Age))

olym_sport %>% 
  group_by(Sport) %>% 
  summarize(count = n(), avg_ht = mean(Height), med_ht = median(Height), sd_ht = sd(Height), IQR_ht = IQR(Height), max_ht = max(Height), min_ht = min(Height))

olym_sport %>% 
  group_by(Sport) %>% 
  summarize(count = n(), avg_wt = mean(Weight), med_wt = median(Weight), sd_wt = sd(Weight), IQR_wt = IQR(Weight), max_wt = max(Weight), min_wt = min(Weight))

head(olym_reg_medals)
```


```{r summaryStats}
summaryStats <- summary(olym_sport)
summaryStats

summary(olym_reg_medals)
```

```{r counting medals}
olym_reg_medals %>%
  group_by(Team) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

olym_reg %>%
  filter(Medal == "Bronze" | Medal == "Silver" | Medal == "Gold") %>% 
  group_by(reg) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

olym_reg %>% 
  filter(is.na(reg))

```

## Filtering by...

### Filtering by Top 5

```{r top5}
olym_top5 <- olym_reg_medals %>%
  filter(Team == "United States" | Team == "Italy" | Team == "Russia" | Team == "Australia" | Team == "Canada")

olym_top5_reg <- olym_reg_medals %>%
  filter(reg == "USA" | Team == "Russia" | Team == "Italy" | Team == "Germany" | Team == "Canada")
```

### Filtering by Sport

```{r bySport}
olym_fence <- olym %>%
  filter(Sport == "Fencing")
summary(olym_fence)

olym %>% 
  filter(Sport == "Tug-Of-War" & Year >= 2000) %>%
  group_by(Team) %>% 
  summarize(count = n()) %>%
  arrange(desc(count))

```

## Grouping by...

### Grouping by Age

```{r byAge}
olym_reg_medals %>% 
  group_by(Age) %>% 
  summarize(medals = n()) %>% 
  arrange(desc(medals))

olym_reg_medals %>% 
  filter(Medal == "Gold") %>%
  filter(!is.na(Age)) %>% 
  summarize(avg_age = mean(Age), median_age = median(Age), sd_age = sd(Age), max_age = max(Age), min_age = min(Age))
```

### Grouping by Medal Type

```{r byMedal}
#Counting number of each medal
olym_reg_medals %>% 
  group_by(Medal) %>%
  summarize(count = n())

#Calculating age stats by medal
olym_reg_medals %>% 
  group_by(Medal) %>% 
  filter(!is.na(Age)) %>% 
  summarize(avgAge = mean(Age), median(Age), sd(Age), min(Age), max(Age))

#Calculating weight stats by medal
olym_reg_medals %>% 
  group_by(Medal) %>%
  filter(!is.na(Height)) %>% 
  summarize(avgHt = mean(Height), median(Height), sd(Height), min(Height), max(Height))

#Calculation height stats by medal
olym_reg_medals %>%
  group_by(Medal) %>%
  filter(!is.na(Weight)) %>% 
  summarize(avgWt = mean(Weight), median(Weight), sd(Weight), min(Weight), max(Weight))

```

- number of bronze, silver, and gold medals awarded are **similar**, but not the same

- average age of bronze, silver, and gold medalists is about equal, at 24 years old

## Plots

```{r plots1}
ggplot(olym_reg_medals, aes(y = Age, color = Medal)) +
  geom_boxplot()
ggplot(olym_reg_medals, aes(x = Season, fill = Medal)) +
  geom_bar()
```

**Boxplot: 25 is the median age across all medals. Outlier ages start around 37. All right skewed.**
**Barplot: Significantly more medals won in summer than winter.**

```{r plots2}
ggplot(olym_top5_reg, aes(x = reg, fill = Medal)) +
  geom_bar()
ggplot(olym_top5_reg, aes(y = Age, color = reg)) +
  geom_boxplot()

ggplot(olym_top5_reg, aes(x = reg, fill = Sport)) +
  geom_bar(position = "fill") +
  labs(y = "proportion", x = "region")
```

**Barplot: Shows that the U.S. has the most medals in all three categories.** 
**Boxplot:SU has the smallest range of ages participating in the olympics, with all being younger than 50. Great Britain has the oldest person to win a medal.**

```{r plots3}
ggplot(olym_top5_reg, aes(x = reg, fill = Medal)) +
  geom_bar(position = "fill")

ggplot(olym_top5_reg, aes(y = Age, color = reg)) +
  geom_boxplot()
```


```{r plots4}
ggplot(data = olym_reg, aes(x = Age)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(vars(is_medalist))

ggplot(data = olym_reg, aes(x = Height, y = Weight, color = Age)) +
  geom_jitter(alpha = 0.75) +
  facet_wrap(vars(is_medalist))
```

```{r plots5}
olym_swim <- olym_reg_medals %>%
  filter(Sport == "Aquatics")

olym_swim_top5 <- olym_swim %>% 
  group_by(reg) %>%
  summarize(medals = n()) %>%
  arrange(desc(medals)) %>% 
  head(n = 5)

ggplot(olym_swim_top5, aes(x = reorder(reg, -medals), y = medals)) +
  geom_col(fill = "deepskyblue") +
  labs(title = "Top 5 Regions in Olympic Aquatics", x = "region", y = "medals won")

olym_swim %>% 
  group_by(reg) %>% 
  mutate(medals = n()) %>% 
  filter(medals >= 15) %>% 
  ungroup() %>% 
  ggplot(aes(x = Age, y = reorder(reg, +medals))) +
  geom_boxplot(color = "darkblue") +
  labs(title = "Distribution of Ages of Olympic Aquatic Medalists", subtitle = "by top 10 regions", x = "age", y = "region")

```

```{r correlation matrix}
numeric_olym_reg <- select_if(olym_reg, is.numeric)

correlation_matrix <- cor(numeric_olym_reg)

correlation_matrix

corrplot(correlation_matrix, method = "color")

heatmap(correlation_matrix)
```


